Starcode: sequence clustering based on all-pairs search

### 2014-11-04 doc/document.tex
Finished discussion.

\section{Discussion and conclusion}
Starocde is a solid algorithm for sequence clustering based on
all-pairs matching. It achieves high precision, and on experimental
datasets it can be faster than popular heuristics. By design,
starcode is tailored to process high throughput sequencing data
on multi-core platforms.
Our benchmark shows that
starcode achieves perfect clustering on short random sequences in less
time than what is considered acceptable for heuristic searches. We also
show that starcode outperforms next generation read mappers in this
context. The software tools used for this benchmark are usually not
used for barcode or random sequence clustering. In this respect,
starcode fills a need arising from the development of barcoding
technologies.

The speed of starcode also makes it useful for other clustering
tasks, such as identifying enriched motifs in microbial genomes and
in experimental data. Here we have given two examples of such
applications. In the first, we recover a known enriched 12-mer in the
genome of \textit{Neisseria meningitidis}. In the second, we recover
the motif of the human RNA binding protein SRSF1 and notice that it
seems to consist of two halves separated by a linker. This hypothesis
is consistent with the fact that SRSF1 binds RNA through two
consecutive RNA Recognition Motifs (RRM) that are known to bind
3-4 nucleotides in a row\citep{pmid23253355}. The Levenshtein distance,
which incorporates insertions and deletions is more likely to capture
bi-partite binding motifs than position weight matrix representations.
The use of a clustering method to tackle this problem is unusual, but it 
it illustrates the potential advantages of distance-based approaches.

The current version of starcode has been primarily optimized for speed.
The memory footprint depends on the number of sequences to cluster
(because the sequences of the input set are loaded in memory as a set of
of tries) and on the mean number of matches per sequence. Every match is
stored in memory until the clustering phase, which may represent a large
overhead if the dataset is dense. As counterintuitive as it may seem,
long queries will usually impose a lower memory footprint because the
matches between the sequences are more sparse. We have shown that the
running time will also be shorter thanks to the lookup table
search (Figure \ref{fig:perf}c).

Perhaps the most surprising element of this study is that an exact
algorithm can compete with extremely fast heuristics. This will no
longer be true when clustering divergent sequences because the Levenshtein
distance will have to be increased, leading to exponentially longer
running times (Figure \ref{fig:perf}b). However, for the imporant
practical case that the divergence is driven by sequencing errors,
starcode illustrates that there is still room for algorithmic innovations
that can outperform heuristics. The idea of the poucet search seems
simple in retrospect, yet it is a powerful way to tap into the data
structuration provided by string sorting. This principle could find some
applications in algorithms used in various fields.

>>>>

\section{Discussion and conclusion}
Starocde is a solid algorithm for sequence clustering based on
all-pairs matching. It achieves high precision, and on experimental
datasets it can be faster than popular heuristics. By design,
starcode is tailored to process high throughput sequencing data
on multi-core platforms with sufficient amount of memory. Due to its
superior precision and faster running time, it fills a gap among
available software, by allowing to take full advantage of middle
to high end hardware.

The speed and precision of starcode also makes it useful for other
clustering tasks, such as identifying enriched motifs in microbial
genomes and in experimental data. Here we have given two examples of such
applications. In the first, we recover a known enriched 12-mer in the
genome of \textit{Neisseria meningitidis}. In the second, we recover
the motif of the human RNA binding protein SRSF1 and notice that it
seems to consist of two halves separated by a linker. This hypothesis
is consistent with the fact that SRSF1 binds RNA through two
consecutive RNA Recognition Motifs (RRM) that are known to bind
3-4 nucleotides in a row\citep{pmid23253355}. The Levenshtein distance,
which incorporates insertions and deletions is more likely to capture
bi-partite binding motifs than position weight matrix representations.
The use of a clustering method to tackle this problem is unusual, but it
it illustrates the potential advantages of distance-based approaches.

One of the reasons why starcode appears to be faster than alternative
tools is that it is designed to cluster relatively similar sequences.
When clustering related but
divergent sequences, the Levenshtein distance will have to be increased,
leading to exponentially longer running times (Figure \ref{fig:perf}b).
However, for the imporant practical case of correcting errors introduced
by sequencing, starcode illustrates that there is still room for
developing algorithms that are both faster and more accurate than
the current state of the art.
